---
title: "Practical Machine Learning Prediction Assignment"
output: 
  html_document:
    keep_md: true
---

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement &#45; 
a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it.

This project uses data from accelerometers on the belt, forearm, arm, and 
dumbell of 6 participants to predict the manner in which they did the exercise. 


##Load data
```{r warning=FALSE}
#Load the neccessary libraries
library(caret)
library(randomForest)

#Download and load train data
trainUrl <-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainFile <- "./data/pml-training.csv"

#If data folder does not exists, create data folder
if (!file.exists("./data")) {
  dir.create("./data")
}

#If data file does not exists, then download the data file
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile)
}

#Load train data
trainData <- read.csv(trainFile)

str(trainData)
```

##Clean data
```{r warning=FALSE}
#Remove NearZeroVariance Variables
nzv <- nearZeroVar(trainData)
trainData <- trainData[, -nzv]

#Identify columns with missing values.
hasMissing <- sapply(trainData, function (x) any(is.na(x) | x == ""))

#Select columns that contains classe, belt, arm, dumbbell and forearm
selectedColumns <- !hasMissing & grepl("classe|belt|[^(fore)]arm|dumbbell|forearm", names(hasMissing))
Predictors <- names(hasMissing)[selectedColumns]
Predictors

trainData <- trainData[,Predictors]
```

Split the cleaned training set into 3/4 for training and 1/4 for validation. 
The validation data set will be used to conduct cross validation.
```{r}
set.seed(123456) # For reproducibile purpose
dp <- createDataPartition(trainData$classe, p=3/4, list=F)
trainSet <- trainData[dp, ]
validationSet <- trainData[-dp, ]
```

##Data Modeling
Selected Random Forest algorithm because it automatically selects important variables and is robust to correlated covariates & outliers. Uses 5-fold cross validation and 250 trees when applying the algorithm.
```{r}
controlRf <- trainControl(method="cv", 5)
fit <- train(classe ~ ., data=trainSet, method="rf", trControl=controlRf, ntree=250)
fit
fit$finalModel
```

Apply the model to the validation set and show the confusion matrix to compare the predicted versus the actual labels.
```{r}
predictRf <- predict(fit, validationSet)

confusionMatrix(validationSet$classe, predictRf)
```

The estimated accuracy of the model is 99.35% and the estimated out-of-sample error is 0.65%.
The out of sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Considering that the test set is a sample size of 20, an accuracy rate above 99% is sufficient and do not require to trying of additional algorithms.

#Prediciting with Testing Dataset
```{r}
#Download and load test data
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testFile  <- "./data/pml-testing.csv"

#If data file does not exists, then download the data file
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile)
}

#Load test data
testData <- read.csv(testFile)

dim(testData)

#Apply the model to the test data set
predictions <- predict(fit,testData)
print(predictions)
```

#Generating Answers Files to Submit for Assignment
Create the answers files for submission
```{r}
#If answers folder does not exists, create answers folder
if (!file.exists("./answers")) {
  dir.create("./answers")
}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./answers/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictions)
```